Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-15 numerical.tex
updating

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield the same results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth} (with $L=1$). We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
in our analysis, the agreement is satisfactory, especially at large
$\khat$.

@@

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(i.e. the branch where $\nu$ increases with $|\omega|$) require larger
vertical domains, since they have increasing number of nodes along 
$z$. The `tail' end with decreasing $\nu$ to the right
of the plot differ significantly between the cases, and are thus
likely associated with boundary conditions.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'.   
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. They typically
have the largest growth rates but require sufficiently large vertical
domain size and their eigenvalues vary with $\zmax$.

Due to their dependence on boundary conditions, we 
do not consider surface modes further in detail. These modes have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode. %  Thus, surface 
% modes may not be the most relevant in realistic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_tags.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$
    \label{compare_modes_iso_kx30}. Examples of surface modes are
    marked by `s'. 
  }
\end{figure}

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$ and $\zmax=7H$. In all 
three cases the number of unstable modes decrease with increasing
$\beta$. The fundamental mode is not the most unstable for 
$\beta=0.01$ as there is a number of higher order modes, but 
their existence depends on $\zmax$.

However, the fundamental mode is dominant for $\beta \geq 
0.03$. Higher order modes, with larger growth rates, are more
effectively stabilized with increasing $\beta$. For $\beta=0.1$, which
is close to $\beta_\mathrm{crit}$, we only find the fundamental
mode. This is consistent with the discussion in 
\S\ref{iso_vsi_beta_crit}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (green triangles) and $\beta=0.1$ (blue
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, our fiducial setup) and $\zmax=7H$
    (bottom). The fundamental mode is marked with `f'.  
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

@@

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. In this case surface modes appear for
$\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the vertical domain size as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z3.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$. Examples of surface modes are marked with `s'. 
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$.  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.

@@

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, the VSI can
operate slightly beyond $\beta_\mathrm{crit}$ (recall that
$\beta_\mathrm{crit}$ was derived by assuming $\khat^2\gg 1$). For
$\khat\geq 50$, we find perturbations have large amplitudes near the
 vertical boundaries, and 
complications to our numerical search of
eigenvalues arise from the emergence of modes associated with boundary
conditions. Nevertheless, growth rates rapidly decrease as $\beta$
approaches and increase beyond $\beta_\mathrm{crit}$.

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stablizes the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that in
\citeauthor{nelson13}). It is interesting to note that there is a
thermal relaxation timescale that maximizes the mode decay rate for
$1\lesssim \khat \lesssim 30$. Here, it is $\beta\simeq1$---$2$ or
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \citeauthor{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

@@

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters
 (Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
 reference point and vary $q\in[-0.2,-1.2]$, 
 $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
 separately.
 % Here, we use a wavenumber $\khat=10$, for which we
 % find growth rates reach zero for the range of disk parameters
 % considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely. 
 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.

>>>>

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield the similar results.) This
is similar to strictly isothermal perturbations in a vertically
isothermal disk considered in \cite{nelson13} and \cite{mcnally14}.

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth}. We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
in our analysis, the agreement is satisfactory.

@@

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(along the analytic lines) require larger vertical domains, since they
have increasing number of nodes along $z$.

The `tail' end with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} is not predicted from
analysis. These modes differ significantly between cases and are thus
likely associated with boundary conditions. % We do not consider them 
% further.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'. Lines
    are computed from the analytic dispersion relation 
    with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. They have the largest growth
rates but require sufficiently large vertical domain size and their
eigenvalues vary with $\zmax$.

Due to their dependence on boundary conditions, we 
do not consider surface modes further in detail. Surface modes have
already been discussed in the aforementioned studies. In fact, we will
see in the next section that surface modes are more effectively
stabilized by finite thermal relaxation than the fundamental mode. %  Thus, surface 
% modes may not be the most relevant in realistic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx30_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_iso_kx10} but for $\khat=30$
    \label{compare_modes_iso_kx30}. Examples of surface modes are
    marked by `s'. 
  }
\end{figure}

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there are modes with $M>0$ that have larger growth 
rates. However, by comparing with results for $\zmax=3H$, we see these
higher order modes are strongly influenced by boundary conditions, and
there is a significant mismatch between numerical and analytical
results for $\zmax=3H$ for $M>1$.

However, the fundamental mode is dominant for $\beta \geq 
0.03$. Higher order modes, with larger growth rates at small $\beta$,
are more effectively stabilized with increasing $\beta$. For
$\beta=0.1$, which is close to $\beta_\mathrm{crit}$, we only find the
fundamental mode. This is consistent with the discussion in 
\S\ref{iso_vsi_beta_crit}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx10_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx10_z7_analytic.ps}
  \caption{Unstable modes with $\khat=10$ and thermal
    relaxation timescales $\beta=0.01$ (black dots), $\beta=0.03$ (red
    crosses), $\beta=0.05$ (green triangles) and $\beta=0.1$ (blue
    crosses); for different vertical domain sizes $\zmax=3H$ (top),
    $\zmax=5H$ (middle, our fiducial setup) and $\zmax=7H$
    (bottom). Lines are computed analytically from
    Eq. \ref{relax_disp} for  $M=0$ (solid, fundamental mode) and
    $M=1,\,2,\,3$ (dotted, dashed, dash-dot, respectively). Along each
    line, $\beta$ increases continously from $0.01$ to $0.1$ from top
    to bottom, and squares mark corresponding $\beta$ values with
    numerical results. 
    \label{compare_modes_cool_kx10} 
  }
\end{figure}

@@

%fundamental mode bcomes the dominant mode as we approach beta
%critical, therefore can use beta crit as criteria

In Fig. \ref{compare_modes_cool_kx30} we plot eigenvalues for unstable
modes with $\khat=30$. For clarity we only plot analytical results for
the fundamental mode, which is again robust against boundary
conditions compared to higher order modes. We find surface modes
appear for $\zmax\geq5H$. However, they are effectively stabilized with
increasing $\beta$. For $\zmax=3H$, the fundamental mode becomes
dominant for $\beta\geq 0.03$, as observed for $\khat=10$. For
$\zmax=5H$ and $\zmax=7H$, we find the emergence modes slightly more
unstable than the fundamental mode even when $\beta=0.1$ (the leftmost
eigenvalues in the lower two panels). However, these modes are clearly associated
with the vertical domain size as they are absent for $\zmax=3H$. 
% Thus, for $\zmax=5H$ and  
% $\zmax=7H$  the fundamental mode is likely dominant for 
% $\beta\geq0.05$

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z3_analytic.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_modes_cool_kx30_z5_analytic.ps}
  \includegraphics[width=\linewidth]{figures/compare_modes_cool_kx30_z7_analytic.ps}
  \caption{Same as Fig. \ref{compare_modes_cool_kx10} but for
    $\khat=30$. Examples of surface modes are marked with `s'. 
    \label{compare_modes_cool_kx30} 
  }
\end{figure}

Fig. \ref{compare_modes_cool_kx10} and \ref{compare_modes_cool_kx30} 
show that the fundamental mode is not sensitive to boundary
conditions. The fundamental mode also becomes dominant as $\beta$
approaches $\beta_\mathrm{crit}$. %beta<beta_crit is necessary and sufficient  
Thus, tracking the fundamental mode is a more robust way of 
assessing stability as $\beta$ is increased from zero, rather than
searching for the most unstable mode, which may correspond to modes
that depend on boundary conditions.

@@

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides
an accurate prediction of the upper limit to the thermal relaxation 
timescale for the fundamental mode. For smaller $\khat$, the VSI can
operate slightly beyond $\beta_\mathrm{crit}$ (recall that
$\beta_\mathrm{crit}$ was derived for $\khat^2\gg 1$). For
$\khat\geq 50$, we find perturbations have large amplitudes near the
 vertical boundaries, and 
complications to our numerical search of
eigenvalues arise from the emergence of modes associated with boundary
conditions. %We generally find boundary effects become more significant
%with increasing $\khat$.  
Nevertheless, growth rates rapidly decrease as $\beta$
approaches and increase beyond $\beta_\mathrm{crit}$.

Fig. \ref{bcrit_compare1} is consistent with numerical simulations
performed by \cite{nelson13} for which a thermal relaxation timescale
$\beta\gtrsim 0.6$ stablized the VSI. This can be anticipated from our
estimate of the critical thermal timescale $\beta_\mathrm{crit}\simeq
0.1$ for our fiducial disk parameters (which is the same as that in
\citeauthor{nelson13}). It is interesting to note that there is a
thermal relaxation timescale that maximizes the mode decay rate for
$1\lesssim \khat \lesssim 30$. Here, it is $\beta\simeq1$---$2$ or
about $0.1$---$0.3$ orbital periods. This is in fact consistent with 
Fig. 12 of \citeauthor{nelson13} where the perturbed kinetic energy in their
numerical simulations were observed to decay most rapidly for a
thermal timescale of $\sim 0.1$ orbits.

@@

 We further test the dependence of the critical thermal relaxation timescale 
 $\beta_\mathrm{crit}$ on disk parameters
 (Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a
 reference point and vary $q\in[-0.2,-1.2]$, 
 $\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
 separately.
 % Here, we use a wavenumber $\khat=10$, for which we
 % find growth rates reach zero for the range of disk parameters
 % considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely. 
 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.
